Why does preregistration increase the persuasiveness of evidence? A Bayesian rationalization.

### 2023-02-13 manuscript.Rmd
minor style changes

Theoretical risk is deeply connected with statistical methods, because it is the inverse of $P(E|\neg{}H)$.
$P(E|\neg{}H)$ is equivalent to the type-I error rate, if you consider the overly simplistic case where the research hypothesis is equal to the statistical alternative hypothesis, because then the nill hypothesis is $\neg{}H$.
Because many researchers are familiar with the type-I error rate, it can be helpful to remember this connection to theoretical risk.
Researchers who choose a smaller type-I error rate can be more sure of their results, if significant, because the theoretical risk is higher.
However, the research hypothesis seldomly equals  the statistical null hypothesis, and therefore, the relation between statistical type-I error rate and theoretical risk should not be overinterpreted.
We argue that theoretical risk (and hence its inverse, $P(E|\neg{}H)$) also encompasses factors outside the statistical realm, most notably the study design and broader analytical strategies.

@@

The argument for a confirmatory research agenda is that by increasing theoretical risk we increase expected epistemic value, i.e., moving to the right on the x-axis in Figure \ref{fig:posterior} increases posterior probability (on the y-axis).
However, if a hypothesis in a certain study has low theoretical risk, researchers cannot do much to change that.
However, studies do not only differ by how high the theoretical risk is but also how certain the recipient is about the theoretical risk.
A study that has a very high theoretical risk, e.g., `r (1-confirmatory_tau)*100`% chance that if the hypothesis is wrong, evidence in its favor will be observed, i.e., alpha error, but has also maximum uncertainty will result in a posterior probability of `r round(confirmatory_uncertain*100)`%, while the same study with maximum certainty will result in `r round(confirmatory_certain*100)`% posterior probability.
The other factors (detectability, prior beliefs, measure of epistemic value) and, therefore, the extend of the benefit very of course with the specifics of the study.
Crucially, even exploratory studies benefit from preregistration, e.g., in this scenario with a $\tau$ = `r some_exploration_tau`, so false positive rate of `r 1-some_exploration_tau`, moving from uncertain to certain increases the posterior by a factor of `r round(some_exploration_certain/some_exploration_uncertain, 2)`.

@@

To summarize, we showed that higher theoretical risk and lower uncertainty about theoretical risk lead to higher expected epistemic value across a variety of measures.
The former result, that increasing theoretical risk leads to higher expected epistemic value, reconstructs the appeal of confirmatory research.
However, theoretical risk is something researcher have only limited control over, e.g., theories are often vague and illdefined, resources are limited, and increasing theoretical risk usually decreases detectability (a special instance of this trade off is the well known tension between alpha error and power).
Before researchers may preregister a confirmatory study with high theoretical risk, they must ensure that they are able to conduct a confirmatory study.
Researchers must develop theories sufficiently and then have the resources to adequately balance detectability against theoretical risk.
A preregistration may at most help to invest some time into developing more specific, hence riskier, implications of a theory, but enabling confirmatory research is a task that requires action from the whole research community.
Incentive structures must be more willing to evaluate not the outcomes of a study but the rigor with which it was conducted [@caganSanFranciscoDeclaration2013; @schonbrodtResponsibleResearchAssessment2022].
Journal editors could encourage theoretical developments and be willing to accept registered reports [@friedLackTheoryBuilding2020; @friedTheoriesModelsWhat2020; @vanrooijTheoryDevelopmentRequires2020; @vanrooijTheoryTestHow2021].
Funding agencies should demand an explicit statement about theoretical risk in relation to detectability and must be willing to provide the necessary resources to reach adequate levels of both [@kooleRewardingReplicationsSure2012].

@@

Focusing on uncertainty reduction also explains two common practices of preregistration that do not align with a confirmatory research agenda.
First, researchers seldomly predict precise numerical outcomes, instead they use preregistrations to describe the process that generates the results.
Precise predictions would have very high theoretical risk (they are likely incorrect if the theory is wrong).
A statistical procedure, may have high or low theoretical risk depending on the specifics of the model used.
Specifying the process, therefore, is in line with the rational we propose here, but is incompatible with strictly confirmatory research.
Second, researchers often have to deviate from the preregistration and make data-dependent decisions outside of the preregistration.
If the only goal of preregistration is to ensure confirmatory research, such changes are not justifiable.

However, under our rational some changes may be justified.
Any change would increase the uncertainty about the theoretical risk and may even decrease the theorethical risk.
The changes still may be worthwhile, if the negative outcomes may be offset by an increase in detectability due to the change.
Consider a preregistration that failed to specify what to do with missing values where the researchers subsequently encounter missings.
In such case, the detectability is zero, since the data cannot be analyzed without a decision about how to handle the missing data.
Any such decision would constitute a deviation from the preregistration.
Furthermore, a reader cannot rule out that the researchers leveraged the decision to decrease theoretical risk, i.e., picking among all options the one that delivers the most beneficial results for the theory.
Whatever decision they make, increased uncertainty about the theoretical risk is inevitable and the expected epistemic value is decreased compared to a world where they anticipated the need to deal with missing data.
However, it is still justified to deviate.
After all they have not anticipated the case and are left with a detectablilty of zero.
Any decision will increase detectability to a non-zero value, likely offsetting the increase in uncertainty.
The researchers also may do their best to argue that the deviation was not motivated by increasing theoretical risk, thereby, decreasing the uncertainty.
Maybe there is one decision that fits well with the theory or with the study design.
Or, if there is no obvious candidate, the researchers could conduct a multiverse analysis of the available options to deal with missings to show the influence of the decision [@steegenIncreasingTransparencyMultiverse2016].

>>>>

One factor that is known without much uncertainty is the contribution of statistical methods to theoretical risk.
Theoretical risk is deeply connected with statistical methods, because it is the inverse of $P(E|\neg{}H)$.
$P(E|\neg{}H)$ is equivalent to the type-I error rate, if you consider the overly simplistic case where the research hypothesis is equal to the statistical alternative hypothesis, because then the nill hypothesis is $\neg{}H$.
Because many researchers are familiar with the type-I error rate, it can be helpful to remember this connection to theoretical risk.
Researchers who choose a smaller type-I error rate can be more sure of their results, if significant, because the theoretical risk is higher.
However, the research hypothesis seldomly equals  the statistical null hypothesis, and therefore, the relation between statistical type-I error rate and theoretical risk should not be overinterpreted.
We argue that theoretical risk (and hence its inverse, $P(E|\neg{}H)$) also encompasses factors outside the statistical realm, most notably the study design and broader analytical strategies.

@@

The argument for a confirmatory research agenda is that by increasing theoretical risk we increase expected epistemic value, i.e., moving to the right on the x-axis in Figure \ref{fig:posterior} increases posterior probability (on the y-axis).
However, if a hypothesis in a certain study has low theoretical risk, researchers cannot do much to change that.
However, studies do not only differ by how high the theoretical risk is but also how certain the recipient is about the theoretical risk.
A study that has a very high theoretical risk, e.g., `r (1-confirmatory_tau)*100`% chance that if the hypothesis is wrong, evidence in its favor will be observed, i.e., alpha error, but has also maximum uncertainty will result in a posterior probability of `r round(confirmatory_uncertain*100)`%, while the same study with maximum certainty will result in `r round(confirmatory_certain*100)`% posterior probability.
The other factors (detectability, prior beliefs, measure of epistemic value) and, therefore, the extend of the benefit very of course with the specifics of the study.
Crucially, even studies with some exploratory aspects benefit from preregistration, e.g., in this scenario with a $\tau$ = `r some_exploration_tau`, so false positive rate of `r 1-some_exploration_tau`, moving from uncertain to certain increases the posterior by a factor of `r round(some_exploration_certain/some_exploration_uncertain, 2)`.

@@

To summarize, we showed that higher theoretical risk and lower uncertainty about theoretical risk lead to higher expected epistemic value across a variety of measures.
The former result, that increasing theoretical risk leads to higher expected epistemic value, reconstructs the appeal of confirmatory research.
However, theoretical risk is something researcher have only limited control over, e.g., theories are often vague and illdefined, resources are limited, and increasing theoretical risk usually decreases detectability (a special instance of this trade off is the well known tension between alpha error and power).
Before researchers may preregister a confirmatory study with high theoretical risk, they must ensure that they are able to conduct a confirmatory study.
Researchers must develop theories sufficiently and then have the resources to adequately balance detectability against theoretical risk.

A preregistration may at most help to invest some time into developing more specific, hence riskier, implications of a theory.
But for a confirmatory science it will not be enough to preregister.
This undertaking requires action from the whole research community [@lishnerConciseSetCore2015].
Incentive structures must be more willing to evaluate not the outcomes of a study but the rigor with which it was conducted [@caganSanFranciscoDeclaration2013; @schonbrodtResponsibleResearchAssessment2022].
Journal editors could encourage theoretical developments and be willing to accept registered reports [@friedLackTheoryBuilding2020; @friedTheoriesModelsWhat2020; @vanrooijTheoryDevelopmentRequires2020; @vanrooijTheoryTestHow2021].
Funding agencies should demand an explicit statement about theoretical risk in relation to detectability and must be willing to provide the necessary resources to reach adequate levels of both [@kooleRewardingReplicationsSure2012].

@@

Focusing on uncertainty reduction also explains two common practices of preregistration that do not align with a confirmatory research agenda.
First, researchers seldomly predict precise numerical outcomes, instead they use preregistrations to describe the process that generates the results.
Precise predictions would have very high theoretical risk (they are likely incorrect if the theory is wrong).
A statistical procedure, may have high or low theoretical risk depending on the specifics of the model used.
Specifying the process, therefore, is in line with the rational we propose here, but is incompatible with strictly confirmatory research.

Second, researchers often have to deviate from the preregistration and make data-dependent decisions outside of the preregistration.
If the only goal of preregistration is to ensure confirmatory research, such changes are not justifiable.
However, under our rational some changes may be justified.
Any change would increase the uncertainty about the theoretical risk and may even decrease the theorethical risk.
The changes still may be worthwhile, if the negative outcomes may be offset by an increase in detectability due to the change.
Consider a preregistration that failed to specify what to do with missing values where the researchers subsequently encounter missings.
In such case, the detectability is zero, since the data cannot be analyzed without a decision about how to handle the missing data.
Any such decision would constitute a deviation from the preregistration.
Furthermore, a reader cannot rule out that the researchers leveraged the decision to decrease theoretical risk, i.e., picking among all options the one that delivers the most beneficial results for the theory.
Whatever decision they make, increased uncertainty about the theoretical risk is inevitable and the expected epistemic value is decreased compared to a world where they anticipated the need to deal with missing data.
However, it is still justified to deviate.
After all they have not anticipated the case and are left with a detectablilty of zero.
Any decision will increase detectability to a non-zero value, likely offsetting the increase in uncertainty.
The researchers also may do their best to argue that the deviation was not motivated by increasing theoretical risk, thereby, decreasing the uncertainty.
Maybe there is one decision that fits well with the theory or with the study design.
Or, if there is no obvious candidate, the researchers could conduct a multiverse analysis of the available options to deal with missings to show the influence of the decision [@steegenIncreasingTransparencyMultiverse2016].
